### 2016-02-03 text/Chapter05.tex
Reorg chapter 4 and 5, adding unique typing to chapter 4

% ==================================================================================================
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
% ==================================================================================================

\documentclass[
		twoside,openright,titlepage,numbers=noenddot,headinclude,%1headlines,
                footinclude=true,cleardoublepage=empty,
                BCOR=10mm,paper=a4,fontsize=10pt, % Binding correction, paper type and font size
                ngerman,american, % Languages
                ]{scrreprt}

\input{classicthesis-config}
\usepackage{semantic}
\usepackage{paralist}
\usepackage{etoolbox}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\newcounter{mythmcounter}
\newcounter{mydefcounter}
\newcounter{myexpcounter}
\newtheorem{example}[myexpcounter]{Example}
\newtheorem{remark}[mythmcounter]{Remark}
\newtheorem{definition}[mydefcounter]{Definition}
\newtheorem{theorem}[mythmcounter]{Theorem}
\newtheorem{lemma}[mythmcounter]{Lemma}
\newtheorem{proposition}[mythmcounter]{Proposition}
\newtheorem*{lemma*}{Lemma}

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.25,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\sep}[0]{\; | \;}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

\newcommand{\ctyp}[2]{C^{#1}#2}
\newcommand{\mtyp}[2]{M^{#1}#2}
\newcommand{\subst}[3]{#1[#2 \leftarrow #3]}

% Typing rule and typing statement (single line)
\newcommand{\tyrule}[3]{ \inference[\footnotesize{(\emph{#1})}~~]{#2}{#3} }
\newcommand{\tyruler}[3]{ \inference{#2}{#3}~~\footnotesize{(\emph{#1})} }

% Coeffect algebra
\definecolor{aclr}{rgb}{0.6,0.3,0.0}
\newcommand{\aclrd}[1]{\textcolor{aclr}{#1}}

% Coeffect algebra
\definecolor{cclr}{rgb}{0.0,0.5,0.0}
\newcommand{\cclrd}[1]{\textcolor{cclr}{#1}}

\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }}
\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
\newcommand{\czip}{ \textcolor{cclr}{\wedge} } % only used by flat coeffects
\newcommand{\cleq}{ \textcolor{cclr}{\leq} }
\newcommand{\cgeq}{ \textcolor{cclr}{\geq} }

\newcommand{\atimes}{ \textcolor{aclr}{\times} }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq
\newcommand{\aparstr}{ \textcolor{aclr}{\oplus}} % Should be the same as \cseq

% Unified

\newcommand{\SH}{ {\textcolor{sclr}{S}} }
\newcommand{\SP}{ {\textcolor{sclr}{P}} }
\newcommand{\SHP}{ {\textcolor{sclr}{S\triangleleft P}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
\newcommand{\sflat}{\star}

\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} }
\newcommand{\aweak}{ \alift{\czero} }

\newcommand{\azip}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\apar}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }

\newcommand{\xtimes}{\hat{\times}}

\newcommand{\ctx}{\textit{ctx}}

% Ordinary context where colouring is done by hand
\newcommand{\coctx}[2]{ #1 \,\text{\scriptsize @}\, {{\text{$#2$}}}  }

% Simple coeffect systems in the introduction
\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}

\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}}

\definecolor{sclr}{rgb}{0.2,0,0.8}
\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }

\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
\newcommand{\M}{ {\textcolor{cclr}{\mathcal{M}}} }

\newcommand{\alen}[1]{{\aclrd{\textit{len}(}#1\aclrd{)}}}
\newcommand{\slen}[1]{{\sclrd{\textit{len}(}#1\sclrd{)}}}
\newcommand{\xlen}[1]{{\textit{len}(#1)}}

\newcommand{\reduce}{\longrightarrow}

\newcommand{\restr}[2]{#1|_{#2}}
\newcommand{\narrow}[1]{\hspace{-0.5em} #1 \hspace{-0.65em}~}

\definecolor{todoclr}{rgb}{0.8,0,0.5}
\newcommand{\todo}[1]{\textcolor{todoclr}{ \begin{quotation} \noindent\textbf{TODO:} #1 \end{quotation} }}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\catc}{\mathcal{C}}
\newcommand{\idf}[1]{ {\textnormal{\sffamily id}}_{#1} }

\newcommand{\figcaption}[1]{
  \vspace{0.5em}
  \noindent\makebox[\linewidth]{\rule{\textwidth}{0.5pt}}
  \vspace{-2em}
  \caption{#1}
}
% Half line spacing around align
\newcommand{\zerodisplayskips}{%
  \setlength{\abovedisplayskip}{0.5em}
  \setlength{\belowdisplayskip}{0.5em}
  \setlength{\abovedisplayshortskip}{0.5em}
  \setlength{\belowdisplayshortskip}{0.5em}}
\appto{\normalsize}{\zerodisplayskips}
\appto{\small}{\zerodisplayskips}
\appto{\footnotesize}{\zerodisplayskips}

% half line spacing around compactitem
\setlength{\pltopsep}{0.5em}
\setlength{\plitemsep}{0.3em}

\patchcmd{\thebibliography}{\chapter*}{\section*}{}{}

\begin{document}

\frenchspacing % Reduces space after periods to make text more compact
\raggedbottom % Makes all pages the height of the text on that page
\selectlanguage{american} % Select your default language - e.g. american or ngerman
\pagenumbering{arabic} % Roman page numbering prior to the start of the thesis content (i, ii, iii, etc)
\pagestyle{plain} % Suppress headers for the pre-content pages

\newcommand{\semdeff}[4]{
\begin{array}{rcl}
  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
  #2 & = & #4 \\
\end{array}  
}
\newcommand{\semdefff}[6]{
\begin{array}{rcl}
  #1 & = & #4 \\
  #2 & = & #5 \\ \cline{1-1}\cline{3-3}
  #3 & = & #6 \\
\end{array}  
}
\newcommand{\semdef}[2]{
\begin{array}{rcl}
  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
  #1 & = & #2 \\
\end{array}  
}
% ==================================================================================================

\chapter{Translation semantics}
\renewcommand*\thesection{\arabic{section}}

@@

\begin{itemize}
\item 
  We define the common subset of the target functional programming language. This includes the
  syntax of the language, reduction rules and typing rules, but it does \emph{not} include 
  coeffect-specific definitions. Well-typed programs written using the common subset of the 
  target language do not get stuck (via progress and preservation), but they may reduce to 
  \ident{error}, e.g. when accessing the head of an empty list.

@@

\section{Introduction}

@@

>>>>

% ==================================================================================================

\chapter{Semantics of flat coeffects}
\label{ch:transl}

@@

\begin{itemize}
\item We present the semantics of the calculus in terms of \emph{indexed comonads}, which are a
  generalization of comonads, a category-theoretical dual of monads (Section~\ref{sec:flat-semantics}).
  This provides a deeper insight into how (and why) the calculus works and shows an 
  intriguing link with effects.

\item 
  We define the common subset of the target functional programming language. This includes the
  syntax of the language, reduction rules and typing rules, but it does \emph{not} include 
  coeffect-specific definitions. Well-typed programs written using the common subset of the 
  target language do not get stuck (via progress and preservation), but they may reduce to 
  \ident{error}, e.g. when accessing the head of an empty list.

@@

% ==================================================================================================
%
%    ###
%     #  #    # ##### #####   ####  #####  #    #  ####  ##### #  ####  #    #
%     #  ##   #   #   #    # #    # #    # #    # #    #   #   # #    # ##   #
%     #  # #  #   #   #    # #    # #    # #    # #        #   # #    # # #  #
%     #  #  # #   #   #####  #    # #    # #    # #        #   # #    # #  # #
%     #  #   ##   #   #   #  #    # #    # #    # #    #   #   # #    # #   ##
%    ### #    #   #   #    #  ####  #####   ####   ####    #   #  ####  #    #
%
% ==================================================================================================

\section{Introduction}
\label{sec:transl-intro}

@@

\subsection{Related work}

Wadler and Thiemann \cite{monads-effects-marriage} famously showed a correspondence between effect 
systems to monads (a topic that has been also discussed by Atkey \cite{monads-parameterised-notions}
and recently revisited by Vazou an Leijen \cite{monads-effects-remarriage}). This line of work
relates effectful functions $\tau_1 \xrightarrow{\sigma} \tau_2$ to monadic computations 
$\tau_1 \rightarrow M^\sigma \tau_2$. In this chapter, we show a similar correspondence between 
\emph{coeffect systems} and \emph{comonads}. However, due to the asymmetry of $\lambda$-calculus, 
defining the semantics in terms of comonadic computations is not a simple mechanical dualisation
of the work on effect systems and monads.

The main purpose of the comonadic semantics presented in this chapter is to provide a motivation 
for the typing of the flat coeffect calculus. Our approach is inspired by the work of Uustalu and
Vene \cite{comonads-notions} who present the semantics of contextual computations (mainly for
data-flow) in terms of comonadic functions $C \tau_1 \rightarrow \tau_2$. Our \emph{indexed 
comonads} annotate the structure with information about the required context, \ie~$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$.
This is similar to the recent work on \emph{parameterized monads} by Katsumata \cite{monads-parametric}.

% ==================================================================================================
%                                              
%      #####  #                                 
%        #    #                                 
%        #    # ##    ###    ###   # ##   #   # 
%        #    ##  #  #   #  #   #  ##  #  #   # 
%        #    #   #  #####  #   #  #      #  ## 
%        #    #   #  #      #   #  #       ## # 
%        #    #   #   ###    ###   #          # 
%                                         #   # 
%                                          ###                                                                         
% ==================================================================================================

\section{Categorical motivation}
\label{sec:flat-semantics}

The type system of flat coeffect calculus arises syntactically, as a generalization of the examples 
discussed in Chapter~\ref{ch:applications}, but we can also obtain it by looking at the categorical 
semantics of context-dependent computations. This is a direction that we explore in this section. 
Although the development presented here is interesting in its own, our main focus is \emph{using} 
categorical semantics to motivate and explain the design of flat coeffect calculus.

\subsection{Categorical semantics}

As discussed in Section~\ref{sec:path-sem}, categorical semantics interprets terms as morphisms
in some category. For typed calculi, the semantics defined by $\sem{-}$ usually interprets typing 
judgements $x_1 \!:\! \tau_1 \ldots x_n \!:\! \tau_n \vdash e: \tau$ as morphisms 
$\sem{\tau_1 \times \ldots \times \tau_n} \rightarrow \sem{\tau}$.

As a best known example, Moggi \cite{monad-notions} showed that the semantics of various effectful 
computations can be captured uniformly using (\emph{strong}) \emph{monads}. In that 
approach, computations are interpreted as $\tau_1 \times \ldots \times \tau_n \rightarrow \mtyp{}{\tau}$
for some monad $\mtyp{}{}$. For example, $\mtyp{}{\alpha} = \alpha \cup \{ \bot \}$ models 
partiality (maybe monad), $\mtyp{}{\alpha} = \mathcal{P}(\alpha)$ models non-determinism (list 
monad) and side-effects can be modelled using $\mtyp{}{\alpha} = S \rightarrow (\alpha \times S)$ 
(state monad). Here, the structure of a strong monad provides necessary ``plumbing'' for composing 
monadic computations -- sequential composition and strength for lifting free variables into the
body of computation under a lambda abstraction.

Following similar approach to Moggi, Uustalu and Vene \cite{comonads-notions} showed that 
(\emph{monoidal}) \emph{comonads} uniformly capture the semantics of various kinds of context-dependent 
computations~\cite{comonads-notions}. For example, data-flow computations over non-empty lists
are modelled using the non-empty list comonad $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.

The monadic and comonadic model outlined here represents at most a binary analysis of effects or 
context-dependence. A function $\tau_1 \rightarrow \tau_2$ performs \emph{no} effects (requires no 
context) whereas $\tau_1 \rightarrow \mtyp{}{\tau_2}$ performs \emph{some} effects and
$\ctyp{}{\tau_1} \rightarrow \tau_2$ requires \emph{some} context\footnote{This is an 
over-simplification as we can use \eg~stacks of monad transformers and model functions with
two different effects using $\tau_1 \rightarrow M_1(M_2~\tau_2)$. However, monad transformers 
require defining complex system of lifting to be composable. Consequently, they are usually used
for capturing different kinds of impurities (exceptions, non-determinism, state), but not for
capturing fine-grained properties (\eg~a set of memory regions that may be accessed by a
stateful computation).}.

In the next section, we introduce \emph{indexed comonads}, which provide a more precise analysis 
and let us model computations with context requirements $\cclrd{r}$ as functions 
$\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ using an \emph{indexed comonad} $\ctyp{\cclrd{r}}{}$.

\subsection{Introducing comonads}

In category theory, \emph{comonad} is a dual of \emph{monad}. As already outlined in 
Section~\ref{sec:path-sem}, we get a definition of a comonad by taking a definition of a monad 
and ``reversing the arrows''. More formally, one of the equivalent definitions of comonad 
looks as follows:

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

\noindent
From the functional programming perspective, we can see $\ctyp{}{}$ as a parametric data type such as
\ident{NEList}. The $\ident{counit}$ operations extracts a value $\alpha$ from a value that carries 
additional context $\ctyp{}{\alpha}$. The $\ident{cobind}$ operation turns a context-dependent function 
$\ctyp{}{\alpha} \rightarrow \beta$ into a function that takes a value with context, applies
the context-dependent function to value(s) in the context and then propagates the context.

As mentioned earlier, Uustalu and Vene \cite{comonads-notions} use comonads to model data-flow
computations. They describe infinite (coinductive) streams and non-empty lists as example comonads.

\begin{example}[Non-empty list]
A non-empty list is a recursive data-type defined as $\ident{NEList}\,\alpha = \alpha + (\alpha \times \ident{NEList}\,\alpha)$.
We write \kvd{inl} and \kvd{inr} for constructors of the left and right cases, respectively. The 
type \ident{NEList} forms a comonad together with the following \ident{counit} and \ident{cobind} mappings:
%
\begin{equation*}
\begin{array}{rclll}
\ident{counit}~l &\narrow{=}& h &\quad&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{counit}~l &\narrow{=}& h &&\textnormal{when}~l=\kvd{inr}~(h, t)\\[0.5em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inl}~(f\,l) &&\textnormal{when}~l=\kvd{inl}~h\\[-0.25em]
\ident{cobind}\,f~l &\narrow{=}& \kvd{inr}~(f\,l,\;\ident{cobind}~f~t) &&\textnormal{when}~l=\kvd{inr}~(h, t)
\end{array}
\end{equation*}
\end{example}

\noindent
The \ident{counit} operation returns the head of the non-empty list. Note that it is crucial that
the list is \emph{non-empty}, because we always need to be able to obtain a value. The \ident{cobind}
operation defined here returns a list of the same length as the original where, for each element, the 
function $f$ is applied on a \emph{suffix} list starting from the element. Using a simplified
notation for list, the result of applying \ident{cobind} to a function that sums elements of a
list gives the following behaviour:
%
\begin{equation*}
\ident{cobind}~\ident{sum}~(7,6,5,4,3,2,1,0) = (28,21,15,10,6,3,1,0)
\end{equation*}
%
The fact that the function $f$ is applied to a \emph{suffix} is important in order to satisfy the
\emph{left identity} law, which requires that $\ident{cobind}~\ident{counit}~l = l$.

It is also interesting to examine some data types that do \emph{not} form a comonad. As already
mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. The \ident{Maybe} 
data type defined as $1 + \alpha$ is not a comonad for the same reason. However, if we consider 
flat coeffect calculus for liveness, it appears natural to model computations as functions 
$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalise comonads 
to \emph{indexed comonads}.

\subsection{Generalising to indexed comonads}
\label{sec:flat-semantics-idx}

The flat coeffect algebra includes a monoid $(\C, \cseq, \cunit)$, which defines the behaviour of
sequential composition, where the annotation $\cunit$ represents a variable access. An indexed 
comonad is formed by a data type (object mapping) $\ctyp{\cclrd{r}}{\alpha}$ where the annotation 
$\cclrd{r}$ determines what context is required.

\begin{definition}
Given a monoid $(\C, \cseq, \cunit)$ with binary operator $\cseq$ and unit $\cunit$, an 
\emph{indexed comonad} over a category $\catc$ is a triple 
$(\ctyp{\cclrd{r}{}}, \ident{counit}_{\cunit}, \ident{cobind}_{\cclrd{r}, \cclrd{s}})$ where:

\begin{compactitem}
\item $\ctyp{\cclrd{r}}{}$ for all $\cclrd{r} \in \C$ is a family of object mappings 
\item $\ident{counit}_{\cunit}$ is a mapping $\ctyp{\cunit}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ is a mapping $(\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{\cclrd{r}\cseq\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta$ and $g:\ctyp{\cclrd{s}}{\beta} \rightarrow \gamma$:f
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}_{\cunit, \cclrd{s}}~\ident{counit}_{\cunit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit}_{\cunit} \circ \ident{cobind}_{\cclrd{r}, \cunit}~f &= f
  \\
\tag{\emph{associativity}}
\hspace{-10em}
  \ident{cobind}_{\cclrd{r}\cseq\cclrd{s},\cclrd{t}}~(g \circ \ident{cobind}_{\cclrd{r}, \cclrd{s}}~f) &= 
    (\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g) \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}~f)
\end{align}
\end{definition}

\noindent
Rather than defining a single mapping $\ctyp{}{}$, we are now defining a family of mappings 
$\ctyp{\cclrd{r}}{}$ indexed by a monoid structure. Similarly, the $\ident{cobind}_{\cclrd{r}, \cclrd{s}}$ 
operation is now formed by a \emph{family} of mappings for different pairs of indices 
$\cclrd{r}, \cclrd{s}$. To be fully precise, $\ident{cobind}$ is a family of natural transformations 
and we should include $\alpha, \beta$ as indices, writing $\ident{cobind}_{\cclrd{r},\cclrd{s}}^{\alpha, \beta}$.
For the purpose of this thesis, it is sufficient to omit the superscripts and treat $\ident{cobind}$ just 
as a family of mappings (rather than natural transformations). When this does not introduce ambiguity, 
we also occasionally omit the subscripts.

The $\ident{counit}$ operation is not defined for all $\cclrd{r} \in \C$, but only for 
the unit $\cunit$. We still include the unit as an index writing $\ident{counit}_{\cunit}$, 
but this is merely for symmetry and as a useful reminder to the reader. Crucially, this means that 
the operation is defined only for special contexts.

If we look at the indices in the laws, we can see that the left and right identity 
require $\cunit$ to be the unit of $\cseq$. Similarly, the associativity law implies the 
associativity of the $\cseq$ operator.

\paragraph{Composition.}
The co-Kleisli category that models sequential composition is formed by the unit arrow (provided
by $\ident{counit}$) together with the (associative) composition operation that composes computations 
with contextual requirements as follows:
%
\begin{equation*}
\begin{array}{ccl}
\textnormal{--}\, \hat{\circ} \,\textnormal{--}&\narrow{:}& (\ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2) 
  \rightarrow (\ctyp{\cclrd{s}}{\tau_2} \rightarrow \tau_3) 
  \rightarrow (\ctyp{\cclrd{r} \cseq \cclrd{s}}{\tau_1} \rightarrow \tau_3) \\
g \, \hat{\circ} \, f &\narrow{=}& g \circ (\ident{cobind}_{\cclrd{r}, \cclrd{s}} f)
\end{array}
\end{equation*}
%
The composition $\hat{\circ}$ best expresses the intention of indexed comonads. Given two functions
with contextual requirements $\cclrd{r}$ and $\cclrd{s}$, their composition is a function that 
requires $\cclrd{r}\,\cseq\,\cclrd{s}$. The contextual requirements propagate \emph{backwards} and
are attached to the input of the composed function.

\paragraph{Examples.}

Any comonad can be turned into an indexed comonad using a trivial monoid. However, indexed comonads
are more general and can be used with other data types, including indexed \ident{Maybe}.

\begin{example}[Comonads]
Any comonad $\ctyp{}{}$ is an indexed comonad with an index provided by a trivial monoid $(\{1\},\ast,1)$
where $1\ast 1 = 1$. The mapping $\ctyp{1}{}$ is the mapping $\ctyp{}{}$ of the underlying comonad. The
operations $\ident{counit}_1$ and $\ident{cobind}_{1,1}$ are defined by the operations $\ident{counit}$
and $\ident{cobind}$ of the comonad.
\end{example}

\begin{example}[Indexed option]
The \emph{indexed option comonad} is defined over a monoid $(\{ \ident{L},\ident{D} \}, \sqcup,\ident{L})$ 
where $\sqcup$ is defined as earlier, \ie~$\ident{L} = \cclrd{r} \sqcup \cclrd{s} \Longleftrightarrow \cclrd{r}=\cclrd{s}=\ident{L}$.
Assuming $1$ is the unit type inhabited by $()$, the mappings are defined as follows:
%
\begin{equation*}
\begin{array}{l}
\ctyp{\ident{L}}{\alpha} = \alpha\\
\ctyp{\ident{D}}{\alpha} = 1\\
\\
\ident{counit}_{\ident{L}} : \ctyp{\ident{L}}{\alpha} \rightarrow \alpha \\
\ident{counit}_{\ident{L}}~v = v\\
\end{array}
\qquad
\begin{array}{lcl}
\multicolumn{3}{l}{
  \ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
    \rightarrow (\ctyp{\cclrd{r}\sqcup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta}) }\\
\ident{cobind}_{\ident{L},\ident{L}}~f~x &\narrow{=}& f~x\\
\ident{cobind}_{\ident{L},\ident{D}}~f~() &\narrow{=}& ()\\
\ident{cobind}_{\ident{D},\ident{L}}~f~() &\narrow{=}& f~()\\
\ident{cobind}_{\ident{D},\ident{D}}~f~() &\narrow{=}& ()\\
\end{array}
\end{equation*}
\end{example}

\noindent
The \emph{indexed option comonad} models the semantics of the liveness coeffect system discussed in 
Section~\ref{sec:applications-flat-live}, where $\ctyp{\ident{L}}{\alpha} = \alpha$ models a live context 
and $\ctyp{\ident{D}}{\alpha}=1$ models a dead context which does not contain a value. The \ident{counit}
operation extracts a value from a live context. As in the direct model discussed in Chapter~\ref{ch:appendix},
the \ident{cobind} operation can be seen as an implementation of dead code elimination. The definition 
only evaluates $f$ when the result is marked as live and is thus required, and it only accesses $x$ if 
the function $f$ requires its input.

The indexed family $\ctyp{\cclrd{r}}{}$ in the above example is analogous to the \ident{Maybe}
(or option) data type $\ident{Maybe}\,\alpha = 1 + \alpha$. As mentioned earlier, this type does not 
permit (non-indexed) comonad structure, because $\ident{counit}~()$ is not defined. This is not a 
problem with indexed comonads, because live contexts are distinguished by the (type-level) coeffect 
annotation and \ident{counit} only needs to be defined on live contexts.

\begin{example}[Indexed product]
The semantics of implicit parameters is modelled by an indexed product comonad. We use a monoid
$(\mathcal{P}(\ident{Id}), \cup, \emptyset)$ where \ident{Id} is the set of (implicit parameter) names.
As previously, all parameters have the type $\rho$. The data type $\ctyp{\cclrd{r}}{\alpha}
= \alpha \times (\cclrd{r} \rightarrow \rho)$ represents a value $\alpha$ together with a function that
associates a parameter value $\rho$ with every implicit parameter name in $\cclrd{r} \subseteq \ident{Id}$.
The cobind and counit operations are defined as:
%
\begin{equation*}
\begin{array}{l}
\ident{counit}_{\emptyset} : \ctyp{\emptyset}{\alpha} \rightarrow \alpha \\[-0.25em]
\ident{counit}_{\emptyset}~(a, g) = a\\[0.5em]
\end{array}
\quad
\begin{array}{l}
\ident{cobind}_{\cclrd{r},\cclrd{s}} ~:~ (\ctyp{\cclrd{r}}{\alpha} \rightarrow \beta) 
    \rightarrow (\ctyp{\cclrd{r}\cup\cclrd{s}}{\alpha} \rightarrow \ctyp{\cclrd{s}}{\beta})\\[-0.25em]
\ident{cobind}_{\cclrd{r},\cclrd{s}}~f~(a,g) = (f(a,\restr{g}{\cclrd{r}}), \restr{g}{\cclrd{s}})\\[0.5em]
\end{array}
\end{equation*}
\end{example}

\noindent
In the definition, we use the notation $(a, g)$ for a pair containing the value of type $\alpha$
together with $g$, which is a function $\cclrd{r} \rightarrow \rho$. The \ident{counit} operation
takes a value and a function (with empty set as a domain), ignores the function and extracts the
value. The \ident{cobind} operation uses the restriction operation $\restr{g}{\cclrd{r}}$ to 
restrict the domain of $g$ to implicit parameters $\cclrd{r}$ and $\cclrd{s}$ in oder to get
implicit parameters required by the argument of $f$ and by the resulting computation, respectively
(\ie~semantically, it \emph{splits} the available context capabilities). The function $g$ passed 
to \ident{cobind} is initially defined on $\cclrd{r} \cup \cclrd{s}$ and so the restriction is 
valid in both cases.

The structure of \emph{indexed comonads} is sufficient to model sequential composition of 
computations that use a single variable (as discussed in Section~\ref{sec:path-sem}). To model 
full $\lambda$-calculus with lambda abstraction and multiple-variable contexts, we need 
additional operations introduced in the next section.

\subsection{Flat indexed comonads}
\label{sec:flat-semantics-monoidal}

Because of the asymmetry of $\lambda$-calculus (discussed in Section~\ref{sec:applications-structure}),
the duality between monads and comonads can no longer help us with defining the additional structure
required to model full $\lambda$-calculus. In comonadic computations, additional information is attached
to the context. In application and lambda abstraction, the context is propagated differently than 
in effectful computations.

To model the effectful $\lambda$-calculus, Moggi~\cite{monad-notions} requires a \emph{strong} monad which
has an additional operation $\ident{strength} : \alpha\times\mtyp{}{\beta} \rightarrow \mtyp{}{(\alpha\times\beta)}$.
This allows lifting of free variables into an effectful computation. In Haskell, strength can be expressed
in the host language and so it is implicit.

To model $\lambda$-calculus with contextual properties, Uustalu and Vene \cite{comonads-notions} 
require \emph{lax semi-monoidal} comonad. This structure requires an additional monoidal operation:
%
\begin{equation*}
\ident{m} : \ctyp{}{\alpha} \times \ctyp{}{\beta} \rightarrow \ctyp{}{(\alpha \times \beta)}
\end{equation*}
%
The \ident{m} operation is needed in the semantics of lambda abstraction. It represents merging of 
contexts and is used to merge the context of the declaration site (containing free variables)
and the call site (containing bound variable). For example, for implicit parameters, this combines
the additional parameters defined in the two contexts.

The semantics of flat coeffect calculus requires operations for \emph{merging}, but also for
\emph{splitting} of contexts. In addition, we also need a lifting operation (similar to $\iota$ 
from Definition~\ref{def:flat-family}) to model sub-coeffecting.

\begin{definition}
Given a flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$,
a \emph{flat indexed comonad} is an indexed comonad over the monoid $(\C, \cseq, \cunit)$
equipped with families of operations $\ident{merge}_{\cclrd{r},\cclrd{s}}$, $\ident{split}_{\cclrd{r},\cclrd{s}}$ 
and $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ where:
%
\begin{compactitem}
\item $\ident{merge}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
  $\ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta} \rightarrow \ctyp{\cclrd{r}\czip\cclrd{s}}{(\alpha \times \beta)}$
\item $\ident{split}_{\cclrd{r},\cclrd{s}}$ is a family of mappings
  $\ctyp{\cclrd{r}\cpar\cclrd{s}}{(\alpha \times \beta)} \rightarrow \ctyp{\cclrd{r}}{\alpha} \times \ctyp{\cclrd{s}}{\beta}$
\item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
  $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{compactitem}
\end{definition}

\noindent
The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a 
single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper 
bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful 
as an informal analogy.}, elucidating the fact that merging may result in the loss of some parts of 
the contexts $\cclrd{r}$ and $\cclrd{s}$. We look at examples of this operation in the next section.

The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
into two separate values. Note that this does not simply duplicate the value, because the additional
context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide 
\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often 
the \emph{least upper-bound}\footnotemark[1].

Finally, $\ident{lift}_{\cclrd{r'}, \cclrd{r}}$ is a family of operations that ``forget'' some part of
a context. This models the sub-coeffecting operation and lets us, for example, forget some of the
available implicit parameters, or turn a live context (containing a value) into a dead context (empty).
